Gas exchange in large rivers influenced by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite

### 2021-09-14 manuscript_v5.Rmd
big ms_v5 push

---
title: "Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
author: 'CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan'
date: "`r Sys.Date()`"
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Key Points
- River channel efficiency effec the relationship between gas exchange and form-drag dissipation of turbulence
- Large, less efficient channel geometry necessitates that near-surface turbulence is dominated by form-drag rather than bed friction
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data

##Keywords
gas exchange, fluvial geomorphology, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin

##Abstract
*AGU Advances. So 250 words here* <br>
*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style. Might be a bold organization but we'll see*

##Plain Language Summary
*Necessary for AGU advances (200 words)*

##1 Introduction
Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used in this study.

@@

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. Even more broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020]. Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g. @zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010]. This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence and $v$ is the kinematic viscosity and *Sc* is the Schmidt number.
$$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$

While this model works reasonably well in non-fluvial environments where turbulence production is largely confined to bottom shear or wind shear [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems. First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a]. Additionally, 'form-drag shear' via river channel banks, meanders, bars, etc. further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999]. Those authors proposed modeling fluvial *k* via 'form-drag dissipation' $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation. Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it is not synonmous with equation 2 as $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate. Likewise, $k\sim\epsilon_D$ usually has a slope much steeper than 1/4.

Relatively less attention has been paid to gas exchange in very large rivers, which is often conceptualized as a hybrid condition of both fluvial and non-fluvial hydraulic properties. To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be influenced by wind shear, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021]. These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 + $\epsilon$ modeled via the log-law-of-the-wall is valid in their simulated rivers and that *k*~streamflow relationships breakdown in large rivers. However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for why this happens.

Further, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a]. It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

@@

In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change in large rivers and how does this relate to one of the classical model's of gas evasion in aquatic systems (equation 2)? We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations. The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.

##2 Data
Numerous datasets were used in this study. Please see Figure 1 for a map of the approximate locations for the data used in this study. We also provide a flowchart detailing the entire study as Figure S1.

![Figure 1: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced. Not mapped here are over 530,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al. 2019) and over 700 gas exchange velocity measurements (Ulseth et al. 2019).](cache/maps/map_fin.jpg)

Gas exchange theory (section 2) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States. This is to our knowledge the largest collection of field measurements of river channel hydraulics. Ultimately, this dataset was filtered down to 171,553 discrete measurements (see Text S1 for our filtering protocol).

BIKER validation (section 4) was performed on 47 SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation (see Appendix A for all variable descriptions and notation used in this study). These are two models for the Saint Lawrence River and one for the Tanana River.

@@

For the $CO_2$ evasion and carbon efflux calculations (section 4), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.

##3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century. Turbulence in river flows is usually generated by shear forces within the river channel (wind shear's influence is usually limited to only the absolute widest rivers). This within-bank shear comes from the riverbed, as well as 'form-drag' via riverbanks, meanders, bars, suspended sediment, etc. Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.

'Form-drag' shear exists at the depth-scale and therefore is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999]. This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, the depth-scale will decrease and form-drag will compose a progressively larger component of near surface dissipation. This concept is illustrated graphically in Figure 2 and represents the geomorphic concept of 'river channel efficiency': all else being constant, a less efficient channel (wide/shallow) will have a longer wetted perimeter and therefore a larger shear surface area than the same flow through a narrow/deep channel [@chowOpenChannelHydraulics1959a]. This means that the inefficient channel will 1) exhibit more form-drag and thus turbulence, and 2) turbulence generated via form-drag will more readily reach the surface. Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$. An efficient channel will have a low $\frac{R_h}{H}$ ratio while an inefficient channel will have a high $\frac{R_h}{H}$ ratio that approaches 1.

![Figure 2: Conceptual model relating river channel efficiency to the turbulent dissipation rate. All else being constant, a less efficient channel shape (wide/shallow) will 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag will more readily reach the surface.](cache/k600_theory/conceptual_model.jpg)

This all suggests that, for incredibly inefficient river channels, $\epsilon_S \approx \epsilon_D$. Put another way, the channel produces so much turbulence from form-drag, and the flow is so shallow relative to its width, that depth-scale dissipation is likely synonmous with near-surface dissipation. This would mean that equation 2 could be equivalently written as $k\sim Sc^{-n}(v\epsilon_D)^{1/4}$ when a channel is sufficiently inefficient. To explore when, if at all, this point is reached we fit models using 701 river $k_{600}$ measurements from @ulsethDistinctAirWater2019a. We defined 'channel inefficiency' using minimum $\frac{R_h}{H}$ thresholds from 0.75 to 0.995 and then we fit the two models outlined in equations 3 and 4. Because all of the data are normalized to Schmidt numbers of 600, we can convienantly fold the viscosity and Schmidt numbers into the model coefficients. The goal of this analysis is to find at what $\frac{R_h}{H}$ threshold does 1) equation 3's b parameter approximate 1/4 and 2) equation 4's goodness-of-fit (via the coefficient of determination $r^2$) reach a suitably high value.
$$\mathbf{(3)} k_{600}=a(\epsilon_D)^{b}$$
$$\mathbf{(4)} k_{600}=\beta(\epsilon_S)^{1/4}$$

The results of this analysis are plotted in Figure 3a, along with channel geometry schematics to highlight the progressive shift from efficient to inefficient river channels. Equation 3's b parameter only begins to approximate 1/4 when $R_h$ is within 1% of the mean flow depth. At this point, the model's goodness-of-fit also reaches a reasonable value of 0.68. This indicates that a river is sufficiently inefficient when its hyradulic radius approximates the mean flow depth, and at that point $\epsilon_S \approx \epsilon_D$. Figures 3b-4d plot the model fits for this final scenario where $\frac{R_h}{H} >= 0.995$ As Figure 3c confirms, the small-eddy model is viable when a channel is so inefficient that dissipation via form-drag approximates the near-surface dissipation rate in equation 2. Figure 4d further confirms this, where the slope of the generalized model (equation 3) equals `r round(models_ulseth[models_ulseth$name == 'Rh=H',]$slope, 2)` which is nearly equivalent to 1/4.

@@

Finally, becuase $\epsilon_D$ refers to a depth-scale quantity, we tested whether a more parsimonious model is possible by removing the 'middle-man', i.e. the Kolmogorov velocity scale, and directly predicting *k* from depth-scale turbulence. Figure 3b plots fitted model results for the $k=\beta_1U_*$ model. For this data, $U_*$ explains 70% of variation in $k_{600}$ and, as expected, the fitted model is nearly identical to Figure 4c in scaling dynamics (as confirmed visually and via the goodness-of-fit). This all indicates that depth-scale turbulent production/dissipation dominates the surface dissipation signal in maximally inefficient river channels. These findings build off of @ulsethDistinctAirWater2019a's intial result that channel slope exerts a drammatic influence on *k* scaling dynamics, further suggesting that fluvial *k* is as much a function of hydraulic geometry as the turbulence properties of the flow.

>>>>

---
title: "Gas exchange in large rivers influenced by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document: 
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Key Points
- River channel efficiency effects the relationship between gas exchange and dissipation of turbulence
- Wide, less "efficient" river channels exert greater influence on near-surface turbulence than narrower, more efficient ones
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data

## Keywords
gas exchange, fluvial geomorphology, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin

## Abstract
*AGU Advances. So 250 words here* <br>
*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style.*

## Plain Language Summary
*Necessary for AGU advances (200 words)*

## 1 Introduction
Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used in this study.

@@

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. More broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020]. Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010]. This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity and *Sc* is the Schmidt number. Note that we directly measuring $\epsilon_S$ is non-trivial and though a frequent model is to impose the log-law-of-the-wall to scale bottom shear to the surface, this necessiarly assumes that bed shear alone effects turbulence which is often not the case.
$$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$

While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear or wind shear [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems. First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a]. Additionally, 'form-drag shear' via river channel banks, meanders, bars, etc. further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999]. Those authors proposed modeling fluvial *k* via form-drag dissipation $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation. Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it should be stressed that $\epsilon_S \neq \epsilon_D$. This is because $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate. Consequently, the, $k\sim\epsilon_D$ relationship usually has a slope much steeper than the 1/4 in equation 2.

Compared to small streams, relatively little attention has been paid to gas exchange in very large rivers. Large rivers are often conceptualized as a hybrid condition of both fluvial and non-fluvial *k* dynamics. To date, the handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021]. This is also presumably limited to forested environments that provide shelter from wind.

These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets. They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers. However, they did not directly compare against *k* measurements made in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for their results.

Finally, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a]. It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

@@

In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change along the stream-to-river continumn and how does this relate to one of the classical models of gas evasion in aquatic systems (equation 2)? We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations. The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.

## 2 Data
Numerous datasets were used in this study. Please see Figure 1 for a map of the approximate locations for the data used in this study. We also provide a flowchart detailing the entire study as Figure S1.

![Figure 1: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced. Not mapped here are over 170,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al. 2019) and over 700 gas exchange velocity measurements (Ulseth et al. 2019).](cache/maps/map_fin.jpg)

Gas exchange theory (section 3) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States. This is to our knowledge the largest collection of field measurements of river channel hydraulics. Ultimately, this dataset was filtered down to 171,553 discrete measurements for this study (see Text S1 for our filtering protocol).

BIKER validation (section 4) was performed on 47 SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation and is explained in section 4. These are two models for the Saint Lawrence River and one for the Tanana River.

@@

For the $CO_2$ evasion and carbon efflux calculations (section 5), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

## 3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century. As noted in the introduction, turbulence in river flows is usually generated by shear forces at the riverbed, but also across the entire flow depth via 'form-drag' from riverbanks, meanders, bars, suspended sediment, etc [@moogStreamReaerationNonuniform1999]. This concept is synonymous with the geomorphic idea of available stream power, or the total energy dissipated from a river reach via the liquid's potential energy along its slope [originally derived from first-principles by @bagnoldSedimentDischargeStream1960; @bagnoldApproachSedimentTransport1966]. Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ (simply the available stream power normalized by the water mass) is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.

While form-drag dissipation is effectively an alternative representation of reach-scale stream power, little attention has been paid to the resulting scale-dependecy of $\epsilon_D$. Available stream power, and thus 'form-drag' shear, exists at the depth-scale and is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999]. This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, its depth will decrease and its wetted perimeter (and therefore total frictional resistance) will increase. This concept is illustrated graphically in Figure 2 and relates the geomorphic concept of 'river channel efficiency' to turbulence: all else being constant, a less efficient channel (wide/shallow) will have greater frictional resistance and therefore form-drag (Chow, 1959). As shown via the orange arrows in Figure 2, the inefficient channel should 1) produce more turbulence, and 2) turbulence generated via form-drag should more readily reach the surface (the red arrows in Figure 2). Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$ (channel area divided by the width). An efficient channel will have a low $\frac{R_h}{H}$ ratio while an inefficient channel will have a high $\frac{R_h}{H}$ ratio that approaches 1.

![Figure 2: Conceptual model relating river channel efficiency to the turbulent dissipation rate. All else being constant, a less efficient channel shape (wide/shallow) should 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag should more readily reach the surface. This is illustrated by the orange arrows, where the inefficient channel arrows more frequently reach the surface (and influence surface turbulence- red arrows).](cache/k600_theory/conceptual_model.jpg)

This all suggests that, for incredibly inefficient river channels, $\epsilon_S \approx \epsilon_D$. The channel produces so much turbulence from form-drag, and the flow is so shallow relative to its width, that depth-scale dissipation should be nearly synonymous with near-surface dissipation and equation 2 could be equivalently written as $k\sim Sc^{-n}(v\epsilon_D)^{1/4}$. To explore when (if at all) this occurs in rivers, we fit models using 701 river $k_{600}$ measurements from @ulsethDistinctAirWater2019a. First, we defined 'channel inefficiency' using minimum $\frac{R_h}{H}$ thresholds ranging from 0.75 to 0.995 and then we fit the two models outlined in equations 3 and 4 (because all of the data are normalized to Schmidt numbers of 600, we convienantly fold the viscosity and Schmidt numbers into the model coefficients). The goal of this analysis is to find at what $\frac{R_h}{H}$ threshold does 1) equation 3's b parameter approximate 1/4 and thus mirror equation 2 and 2) equation 4's goodness-of-fit (via the coefficient of determination $r^2$) reach a suitably high value that it is deemed physically valid.
$$\mathbf{(3)} k_{600}=a(\epsilon_D)^{b}$$
$$\mathbf{(4)} k_{600}=\beta(\epsilon_D)^{1/4}$$

The results of this analysis are plotted in Figure 3a, along with channel geometry schematics to highlight the progressive shift from efficient to inefficient river channels. Equation 3's b parameter only begins to approximate 1/4 when $R_h$ is within 1% of the mean flow depth. At this point, the model's goodness-of-fit also reaches a reasonable value of `round(models_eD4[models_eD4$name == 'Rh=H',]$r2,2)`. This indicates that a river is sufficiently inefficient when its hyradulic radius approximates the mean flow depth, and at that point $\epsilon_S \approx \epsilon_D$. Figures 3b and 3c plot the model fits for this final scenario where $\frac{R_h}{H} >= 0.995$ As Figure 3c confirms, the small-eddy model is viable when a channel is so inefficient that dissipation via form-drag approximates the near-surface dissipation rate in equation 2. Figure 3b further confirms this, where the slope of the generalized model (equation 3) equals `r round(models_ulseth[models_ulseth$name == 'Rh=H',]$slope, 2)` which is nearly equivalent to 1/4.

@@

Finally, becuase $\epsilon_D$ refers to a depth-scale quantity, we tested whether a more parsimonious model is possible by removing the 'middle-man', i.e. the Kolmogorov velocity scale that the small-eddy model is based on [@lamontEddyCellModel1970], and directly predicting *k* from depth-scale turbulence. Figure 3d plots the fitted $k=\beta_1U_*$ model. For this data, $U_*$ explains 70% of variation in $k_{600}$ and, as expected, the fitted model is nearly identical to Figure 3c in scaling dynamics and variance explained. This all indicates that depth-scale turbulent production/dissipation dominates the surface dissipation signal in extremely inefficient river channels. These findings build off of @ulsethDistinctAirWater2019a's intial result that channel slope exerts a drammatic influence on *k* scaling dynamics, further suggesting that fluvial *k* is as much a function of hydraulic geometry as the turbulence properties of the flow.

With all of that said, how often is a river channel so inefficient that $R_h \approx H$? To answer this question, we join the 701 measurements from @ulsethDistinctAirWater2019a to 171,553 discrete river hydraulics measurements from across the continental United States [@brinkerhoffReconcilingAtaStationAtManyStations2019]. We find that `r round(percs_overall[percs_overall$flag_depth == 'Rh=H',]$perc,0)`% of hydraulic radii are within 5% of their mean flow depth. Because these measurements were made exclusively at streamgauges where rivers are somewhat channelized and width variability is decreased, we opted for a slightly more liberal ratio of 5%. Further, because larg rivers are more likely to have inefficient channels, we identify the measurements that are SWOT-observable, i.e. width > 100m. We find that while `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[1]`% of non-SWOT observable rivers are inefficient, `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[2]`% of SWOT-observable rivers are inefficient. This significant difference, and near universality of SWOT-observable rivers being inefficient, suggests that our theoretical findings in Figure 3 should reasonably hold in nearly all rivers that SWOT will observe. This opens the door for exploiting our 'inefficient river k model' for remote sensing *k* via the SWOT satellite.
